---
title: "Is there an effect of exposure in each group?"
author: "Lucas Moitinho-Silva"
date: "`r Sys.Date()`"
output:
   BiocStyle::html_document:
      toc: true
      code: hide
      df_print: paged
      self_contained: true
      code_download: true
      highlight: tango
#bibliography: mylib.bib
editor_options:
  chunk_output_type: console
params:
  FIGPATH: "figures/"
  d.out: "./"
---


# Objective

Investigate whether there was a change between visits *within* each group. To recap, we have two main groups based on treatment:


1. Control
2. Treatment

We also have two Visits:

1. Before (visit 3)
3. After (visit 5)

I will investigate changes at the community level.

# Preparations

## Set global options

```{r style, echo = T, results="asis", cache=FALSE, message = F}
#v.01
# Set knit global options
library("knitr")
options(digits = 2, width = 80)
golden_ratio <- (1 + sqrt(5)) / 2
opts_chunk$set(echo = TRUE,
               tidy = FALSE,
               include = TRUE,
               fig.path = params$FIGPATH,
               dev=c("png",'pdf'),
               fig.height = 6,
               fig.width = 5 * golden_ratio,
               comment = '  ',
               dpi = 300,
               cache = FALSE,
               warning = F,
               message = F)
library("rmarkdown")

# Pretty outputs
library("BiocStyle")
library("ggthemes") # add ggplot2 themes
library("ggplot2")
theme_set(theme_few(base_size = 14))

# Set seed for reproducibility
set.seed(13)
# Color blind friendly pallet
cbp1 <- c("#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

scale_fill_discrete <- function(...) {
  scale_fill_manual(..., values = cbp1)
}
scale_colour_discrete <- function(...) {
  scale_colour_manual(..., values = cbp1)
}

# Set output directory
d.out <- params$d.out
rm(params)
```


## Load libraries for the session

```{r}
library(ggplot2)
library(tidyverse)
library(broom)
library(reshape2)
library(RColorBrewer)
library(ggrepel)
library(vegan)
library(spaa)
```

## Import data
```{r}
data.folder <- "../data"

micro <- "micro.r.rds" %>% 
  paste0(data.folder, "/", .) %>% 
  readRDS()

meta <- "meta.rds" %>% 
  paste0(data.folder, "/", .) %>% 
  readRDS() %>% 
  mutate(Treatment = factor(Treatment, levels = c("Control",
                                          "Treatment")))

tax <- "tax.rds" %>% 
  paste0(data.folder, "/", .) %>% 
  readRDS()
```


# Alpha

## Calculate
```{r}
meta <- diversity(micro, "shannon") %>% data.frame( NGS_ID = names(.),
                                                            Shannon = .) %>% 
  full_join(meta, by = "NGS_ID")
```

## Make tests
```{r}
group <- "Treatment"
df <- meta %>% 
  filter(Treatment  == "Treatment") %>% 
  select(Visit, Ind, Shannon, Treatment) %>% 
  arrange(Visit, Ind)
wilcox.test(Shannon ~ Visit, data = df,  paired = T)

group <- "Control"
df <- meta %>% 
  filter(Treatment  == "Control") %>% 
  select(Visit, Ind, Shannon, Treatment) %>% 
  arrange(Visit, Ind)

wilcox.test(Shannon ~ Visit, data = df,  paired = T)

```

## Plot

```{r}
to.plot <- meta %>% 
  arrange(Visit, Ind, Treatment) %>%
  mutate(Treatment = factor(Treatment, levels = c("Treatment", "Control")))
p1 <- ggpubr::ggboxplot(to.plot,
                       x = "Visit",
                       y = "Shannon",
                       add = "point",
                       color = "Treatment",
                       facet.by = "Treatment") +
  ggpubr::stat_compare_means(paired = T, label.x = 0.9, label = "p.format") +
   theme(legend.position = "none") 
p1 <- ggpubr::set_palette(p1, palette = c("red", "black")) +
  geom_path(color = "grey70", aes(group = Ind),
            stat = "identity",
            size = 0.5, alpha = 0.9) +
   ggpubr::theme_pubr() +
  theme(legend.position = "none") +
  ggpubr::rotate_x_text()
p1
```


# Beta

## Test effect of visit using adonis2


Design is: ~ Individuals + Visit. I am interested in the effect of visit. Individuals are included in the formula to control for paired samples.
## Function
```{r}
test.adonis <- function(meta, micro, group){
meta.test <- meta %>% 
  filter(Treatment == !!group)

micro.test <- meta.test %>%
  select(NGS_ID) %>% 
  left_join(micro %>% 
              rownames_to_column("NGS_ID"), by = "NGS_ID") %>% 
  column_to_rownames("NGS_ID")

micro.test.dist <-  micro.test %>%
  vegdist("bray")

adonis2(micro.test.dist ~ Ind + Visit, data = meta.test, by = "terms")}
```

### Treatment

```{r}
test.adonis(meta, micro, "Treatment")
```

### Untreat

```{r}
test.adonis(meta, micro, "Control")
```

## Visualize beta diversity

### Calculate MDS
```{r}
micro.dis <- micro %>%
  vegdist("bray")
mds <- micro.dis %>% 
  metaMDS(.,autotransform = F, trymax = 50)
```

### Plot MDS

```{r}
to.plot <- mds$points %>%
  data.frame() %>% 
  rownames_to_column("NGS_ID") %>% 
  inner_join(meta, by = "NGS_ID") %>% 
  mutate(Treatment = factor(Treatment, levels = Treatment %>% 
                              as.character %>% 
                              unique() %>%
                              rev()) )

p0 <- ggplot(to.plot, aes(MDS1, MDS2)) +
  geom_point(aes(color = Visit)) + 
  facet_wrap(~Treatment) +
  stat_ellipse(aes(color = Visit)) +
  geom_path(color = "grey70", aes(group = Ind),
            stat = "identity",
            size = 0.5, alpha = 0.9) +
    ggpubr::theme_pubr() +
  theme(legend.position = "bottom")
p0
```

## Compare pairwise beta diversity

Now, I want to test if the dissimilarities between visits are different between groups.

## Get pairwise dissimilarities
```{r}
pair <- dist2list(micro.dis) %>% 
  left_join(meta %>% select(NGS_ID, Ind, Treatment), by = c("col" = "NGS_ID")) %>% 
  left_join(meta %>% select(NGS_ID, Ind), by = c("row" = "NGS_ID")) %>% 
  #remove instance of the same sample
  filter(row != col) %>% 
  # Keep only distances between the same individual
  filter(Ind.x == Ind.y) %>% 
  #remove duplicated data
  select(Ind.x, Treatment, value) %>% 
  unique() %>% 
  rename("Ind" = "Ind.x") %>% 
  arrange(Treatment) %>% 
  mutate(Treatment = factor(Treatment, levels = Treatment %>% 
                              as.character %>% 
                              unique() %>%
                              rev()) )
```

### Test and visualize

```{r}
p <- ggpubr::ggboxplot(pair, x = "Treatment", y = "value", add = "jitter", color = "Treatment")+
  ggpubr::stat_compare_means(paired = F, label.x = 0.9, label = "p.format") +
  labs(y = "Bray-Curtis dissimilarity\n(intra-individual)") +
  theme(legend.position = "none") +
  ggpubr::rotate_x_text()
p <- ggpubr::set_palette(p, palette = c("red", "black"))
p

wilcox.test(value ~ Treatment, data = pair)
```

# Combine figure

```{r}
comb <- ggpubr::ggarrange(p1, p0, p, nrow = 1, widths = c(1,2,1), labels = "auto")
comb
```

# Save

```{r}
paste0(d.out, "/", "alpha.beta.rds") %>% 
  saveRDS(comb,.)
```


# Session information

```{r}
sessionInfo()
```
