---
title: "Is there an effect of exposure in each group?"
author: "Lucas Moitinho-Silva"
date: "`r Sys.Date()`"
output:
   BiocStyle::html_document:
      toc: true
      code: hide
      df_print: paged
      self_contained: true
      code_download: true
      highlight: tango
#bibliography: mylib.bib
editor_options:
  chunk_output_type: console
params:
  FIGPATH: "figures/"
  d.out: "./"
---


# Objective

Investigate whether there was a change between visits *within* each group. To recap, we have two main groups based on treatment:


1. Control
2. Treatment

We also have two Visits:

1. Before (visit 3)
3. After (visit 5)

I will investigate changes at the ASV level. To reduce multiple test issues, I will only test the differences in the 5 most abundant Staphylococcus ASVs.


# Preparations

## Set global options

```{r style, echo = T, results="asis", cache=FALSE, message = F}
#v.01
# Set knit global options
library("knitr")
options(digits = 2, width = 80)
golden_ratio <- (1 + sqrt(5)) / 2
opts_chunk$set(echo = TRUE,
               tidy = FALSE,
               include = TRUE,
               fig.path = params$FIGPATH,
               dev=c("png",'pdf'),
               fig.height = 6,
               fig.width = 5 * golden_ratio,
               comment = '  ',
               dpi = 300,
               cache = FALSE,
               warning = F,
               message = F)
library("rmarkdown")

# Pretty outputs
library("BiocStyle")
library("ggthemes") # add ggplot2 themes
library("ggplot2")
theme_set(theme_few(base_size = 14))

# Set seed for reproducibility
set.seed(13)
# Color blind friendly pallet
cbp1 <- c("#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

scale_fill_discrete <- function(...) {
  scale_fill_manual(..., values = cbp1)
}
scale_colour_discrete <- function(...) {
  scale_colour_manual(..., values = cbp1)
}

# Set output directory
d.out <- params$d.out
rm(params)
```


## Load libraries for the session

```{r}
library(ggplot2)
library(tidyverse)
library(broom)
library(reshape2)
library(RColorBrewer)
library(ggrepel)
library(scales)
library(DESeq2)
```

## Import data
```{r}
data.folder <- "../data"

micro <- "micro.rds" %>% 
  paste0(data.folder, "/", .) %>% 
  readRDS()

micro.r <- "micro.r.rds" %>% 
  paste0(data.folder, "/", .) %>% 
  readRDS()

meta <- "meta.rds" %>% 
  paste0(data.folder, "/", .) %>% 
  readRDS() %>% 
  mutate(Treatment = factor(Treatment, levels = c("Control",
                                                  "Treatment")))

tax <- "tax.rds" %>% 
  paste0(data.folder, "/", .) %>% 
  readRDS()
```


# Get most abundant ASVs


```{r}
micro.st <- micro.r %>% 
  rownames_to_column("NGS_ID") %>% 
  select(tax$ASV, NGS_ID) %>% 
  melt(variable.name = "ASV", value.name = "Value") %>% 
  left_join(tax %>% select(ASV, Genus, Species), by = "ASV") %>% 
  mutate(Value = (Value * 100)/10000) %>% 
  group_by(Genus, Species, ASV) %>% 
  summarise(rel.abundance = mean(Value),
            rel.abundance.sd = sd(Value)) %>%
  ungroup() %>% 
  mutate(rank = rank(-rel.abundance)) %>% 
  mutate(Species = if_else(str_detect(Species, "unc."),
                           Species,
                           paste(Genus, Species)))


asv.to.keep <- micro.st %>% 
  ungroup %>% 
  filter(Genus == "Staphylococcus") %>% 
  arrange(rank) %>% 
  top_n(5, rel.abundance) %>% 
  pull(ASV)

asv.staphy <- micro.st %>% 
  ungroup %>% 
  filter(Genus == "Staphylococcus") %>% 
  pull(ASV)


ggplot(micro.st, aes(rank, rel.abundance, label = ASV)) +
  geom_point(data = micro.st %>% 
               filter(ASV %in% asv.staphy), aes(color = Species), size =3) +
  ggrepel::geom_text_repel(data = micro.st %>% 
               filter(ASV %in% asv.to.keep)) +
  geom_hline(yintercept = 1) +
  theme(legend.position = "right") +
  geom_point(data = micro.st %>% 
               filter(!ASV %in% asv.staphy), color = "black") +
  labs(x = "Rank", y = "Abundance (%)")

```

# Which percentage the top 5 Staphylococcus ASVs represents of the total sequences? 

```{r}
total.sum <- micro.r %>% sum()
micro.r %>% 
  rownames_to_column("NGS_ID") %>% 
  select(tax$ASV, NGS_ID) %>% 
  melt(variable.name = "ASV", value.name = "Value") %>% 
  left_join(tax %>% select(ASV, Genus, Species), by = "ASV") %>% 
  filter(ASV %in% asv.to.keep) %>% 
  summarize(sum = sum(Value),
            total.sum = total.sum) %>% 
  mutate(percentage = (sum *100)/ total.sum) %>% 
  pull(percentage) %>% 
  print()
```

# Test
## Function

Design is: ~ Individuals + Visit. I am interested in the effect of visit. Individuals are included in the formula to control for paired samples. 

```{r}
test.deseq <- function(micro, meta, asv.to.keep, treatment){
  meta <- meta %>% 
    filter(Treatment == treatment) %>% 
    mutate(Visit = sub(" ", "_", Visit)) %>% 
    mutate(Visit = factor(Visit, levels = c("Day_3","Day_5")))
  micro <- micro %>% 
    rownames_to_column("NGS_ID") %>% 
    right_join(meta %>% select(NGS_ID), by = "NGS_ID") %>% 
    column_to_rownames("NGS_ID")
  
  
  # Transform
  micro.t <- t(micro) %>% 
    data.frame()
  
  # build DESeq object
  dds <- DESeqDataSetFromMatrix(countData = micro.t,
                                colData = meta,
                                design = ~ Ind + Visit)
  #Estimate size 
  dds <- estimateSizeFactors(dds, "poscounts") 
  
  # Keep only most abundant genus
  
  idx <- rownames(micro.t) %in% asv.to.keep
  names(idx) <- rownames(micro.t)
  dds <- dds[idx,]
  
  # Perform test
  dds <- DESeq(dds, reduced = ~ Ind, test = "LRT")
  # Get results
  res <- results(dds, independentFiltering = F,  tidy = T) %>% 
    mutate(Treatment = treatment)
  return(res)
}


```

## Perform test
```{r}
res <- bind_rows(test.deseq(micro, meta, asv.to.keep, "Control"),
                 test.deseq(micro, meta, asv.to.keep, "Treatment")) %>% 
  dplyr::rename("ASV" = "row")
```

## Annotate results
```{r}
res <- res %>% left_join(tax, by = "ASV") %>% select(ASV, Treatment, pvalue, Species, everything())

```

# Visualization

```{r}
micro.st <- micro.r %>% 
  rownames_to_column("NGS_ID") %>% 
  select(tax$ASV, NGS_ID) %>% 
  melt(variable.name = "ASV", value.name = "Value") %>% 
  left_join(meta %>% select(Ind, Visit, Treatment, NGS_ID), by = "NGS_ID") %>% 
  filter(ASV %in% asv.to.keep) %>% 
  left_join(tax %>% select(ASV, Genus, Species), by ="ASV") %>% 
  mutate(Species = if_else(str_detect(Species, "unc."),
                           Species,
                           paste("S.", Species))) %>% 
  mutate(ASV.order = gsub("ASV", "", ASV) %>%  as.numeric) %>% 
  arrange(ASV.order) %>% 
  mutate(ASV = paste0(Species, " ", gsub("ASV", "", ASV))) %>% 
  mutate(ASV = factor(ASV, levels = unique(ASV))) %>% 
  mutate(Visit.plot = if_else(Treatment == "Treatment",
                              paste0("  ", Visit),
                              paste0(" ", Visit)))

```


## Make label

```{r}
label <- res %>% 
  filter(pvalue < 0.05) %>% 
  mutate(label = "*",
         hjustvar = 0,
         vjustvar = 1.1,
         xpos = -Inf,
         ypos = Inf,
         annotateText = paste0(" p: ", round(pvalue, 3))) %>%
  mutate(ASV = paste0(Species, " ", gsub("ASV", "", ASV))) %>% 
  mutate(ASV = factor(ASV, levels = unique(ASV))) %>% 
  mutate(hjustvar = if_else(Treatment == "Treatment",
                            0,
                            1.1),
         xpos = if_else(Treatment == "Treatment",
                            -Inf,
                            Inf))
```
## Plot
```{r, fig.width= 10}
p <- ggplot(micro.st, aes(Visit.plot, Value+0.1)) +
  geom_point(aes(color = Treatment)) +
  facet_wrap(~ ASV, ncol = 5) +
  theme(strip.text.y.right = element_text(angle = 0, hjust = 0)) +
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                labels = trans_format("log10", math_format(10^.x))) +
  geom_path(color = "grey70", aes(group = Ind),
            stat = "identity",
            size = 0.5, alpha = 0.9) +
  scale_color_manual(values = c( "black", "red")) +
    ggpubr::theme_pubr() +
  labs(y = "Counts (total 10K)",  x = "") +
  geom_text(data = label, aes(x = xpos, y = ypos,
                              hjust = hjustvar,
                              vjust = vjustvar,
                              label = annotateText,
                              col = Treatment)) +
  ggpubr::rotate_x_text() +
    theme(legend.position = "bottom")

p
```
# Save output

```{r}
paste0(d.out, "/", "asvs.rds") %>% 
  saveRDS(p,.)

writexl::write_xlsx(res, paste0(d.out, "/", "ASV.results.xlsx"))
```


# Conclusion

# Session information

```{r}
sessionInfo()
```
